Exploratory data structure comparisons: Three new tools based on Principal Component Analysis

### 2018-03-08 article/Old versions/article_v0.01.tex
Redid folder structure + added first print/summary functions

\section{Data example stuff}
\label{sec:dataexample}
We will now turn to a concrete data example in order to illustrate the possibilities of each of the methods presented above \hl{more than just PCADSC?}. We use data from the 2012 version of the European Social Survey (ESS) project, a very large dataset that is freely available online at \url{www.europeansocialsurvey.org}. As the name suggests, the data comes from a survey that was conducted primarily in Europe aiming to collect information about the social conditions of the citizens \hl{reference?}. As with all international (or, simply, multi-center) studies, one might be concerned about whether or not the data from different countries can readily be combined. This is the question we will address using PCADSC in the current section.

@@

\subsection{Data}
\begin{figure}
\center
\includegraphics[scale=1]{surveyTable.png}
\caption{\hl{Make a new table here}}
\label{surveyTable}
\end{figure}
The ESS 2012 data contains a total of 626 variables collected from 54673 citizens in 29 countries. Here, we will only work with a subset of 35 questionnaire items that are all related to psychological well-being. These 35 items can be divided into 6 distinct scales (\hl{REF: ESS6 Topline Results Series 5}), as illustrated in Figure \ref{surveyTable}. We represent each of these scales by a single variable, which is calculated as the average score within the items related to that variable (\hl{ref to someone that says that is sensible?}). For simplicity, we use only complete cases for this construction and thus exclude all participants that did not answer all the 35 questionnaire items used below.

In the following, we only compare two countries, namely Denmark and Bulgaria, which have $n_{BG} = 1798$ and $n_{DK} = 1498$ complete cases in the variables of interest, respectively. For these two countries, the ESS authors \hl{different term?} particularly highlight differences in the relationship between the psychological well-being scales, at least on an nation-aggregated level (\hl{REF: ESS6 Topline Results Series 5}). This might be the result of large, cultural and socio-economic differences between the two countries. Simply put, we do not expect happiness and psychological well-being to be the same phenomena in Bulgaria and Denmark \hl{more here about why, about the cultures?}. Therefore, a successful method for data comparisons should be able to detect these differences by looking at the differences in the interplay between the 6 scales of psychological well-being.

\subsection{PCADSC}
\begin{figure}
\center
\includegraphics[scale = 0.8]{essPCADSC.pdf}
\caption{\hl{something}}
\label{plotESSPCADSC}
\end{figure}

\begin{figure}
\center
\includegraphics[scale = 0.8]{essWallyPCADSC.pdf}
\caption{\hl{something.}}
\label{plotESSPCADSCWally}
\end{figure}

\begin{figure}
\center
\includegraphics[scale = 0.8]{essHairplot.png}
\caption{The blue arrows show the principal components of the first dataset decomposed in the coordinate system of the principal components of the second dataset. And the red arrows the reverse.}
\label{plotESSHairplot}
\end{figure}

\begin{figure}
\center
\includegraphics[scale = 0.8]{essCumVarAgreement.png}
\caption{\hl{something.}}
\label{plotCumVarAgreement}
\end{figure}

Figure \ref{plotESSPCADSC} illustrates the results of conducting PCADSC on the psychological well-being data from the ESS. While the first principal component, which is responsible for explaining 50-60 \% of the variance in the data, is very similar for the two countries, we see quite large differences in the remaining components. In the second component, we see that the two countries disagree in the relative importance of the scales \textit{Community wellbeing} and \textit{Vitality}. In the third and fourth components, general disagreement  is found. All in all, components 2-4, representing almost half of the variability in the data, are not very similar across the two countries. Moreover, the two subsets of the data also differ with respect to how much variance is explained by each component, and the difference is particularly big for the first component. This component has approximately 15 \%  more explanatory power in the Bulgarian subsample than it does in the Danish.

{\color{red}
Figure \ref{plotESSHairplot} shows the hairplot. The blue arrows visualize the decomposition of the principal components for the first dataset in the coordinate system of the second dataset. We see that PC2 also loads on PC3, that PC3 also loads on PC4, and that PC4 also loads on PC2 and PC3. The red arrows visualize the decomposition of the principal components for the second dataset in the coordinate system of the first dataset. We see that PC2 also loads on PC4, that PC3 also loads on PC2 and PC4, and that PC4 also loads on PC3. For PC1, PC5 and PC6 the main difference is in the size of the variation component.
}

In summary, we find that the datasets are fundamentally different and that we should therefore be wary about combining them in a joint analysis, which was also the conclusion of the ESS authors, though based on country-level aggregated statistics. \\

But did we really illustrate a data structure difference due to country differences or did we just illustrate the variability of the results of the PCADSC method? In order to investigate this further, we look at a so-called \textit{Wally plot} (\hl{ref: Claus Ekstrøm}). In this plot, we compare the results of PCADSC conducted with grouping by country with several random, but similar grouping variables. Specifically, we produce 7 PCADSC plots where the country variable was replaced by a randomly generated variable that divides the observations into two groups of the same sizes as the country samples. The results are illustrated in Figure \ref{plotESSPCADSCWally}. Here, we see that the differences in the second component from the original PCADSC results are not matched in any of the randomly grouped PCADSC runs. In fact, the 7 runs are remarkably similar, thereby illustrating that PCADSC seems to be very robust with respect to random groupings: The signal in the data is not blurred by the random subdivisions. When it comes to the differences in the third component for the two groups, we find much larger variability in the 7 random runs. \hl{more comments here... Wait until we are sure exactly what we think about the results and what other PCA-based methods, we will do before/after. Particularly, how do we deal with eigen value differences?}

\section{Discussion}
\label{sec:discussion}
\hl{
\begin{itemize}
\item Generalizing the results to non-numeric variables?
\item Generalizing the results to covariance matrices that are not of full rank?
\item ?
\end{itemize}
}

\section{Concluding Remarks}
\label{sec:conclusion}

>>>>

\section{European differences in psychological well-being: A data example}
\label{sec:dataexample}
We will now turn to a concrete data example in order to illustrate the capabilities of the methods presented above. We use data from the 2012 version of the European Social Survey (ESS) project to investigate inter-country differences in psychological well-being and happiness. This investigation is motivated by an increasingly popular new tendency to publish miscellaneous rankings of countries in fields as different as educational quality (\hl{ref: PISA stuff}), \hl{XXX} and citizen happiness (\hl{ref OECD?}). From a methodological point of view, such international rankings are very concerning, as they rely on the fundamental assumption  that the measured concepts are inherently the same across countries. The PCADSC tools qualify as a suite of methods for exploring the validity of this assumption empirically.

In the rankings of happiness, Denmark has repeatedly been awarded with the title of "happiest country" (\hl{ref?}. However, little critical discussion has been aimed at whether or not this title is in any way meaningful. Does there exist such a thing as a universal, internationally valid concept of happiness? Or do different aspects of psychological well-being or happiness simply do not have the same relative meaning in different cultural and socioeconomic settings? This is in fact a question concerning comparability of data structures. If two countries differ e.g. in terms of how social networks are typically build and structured, with one emphasizing family relations and the other mostly focusing on other social relations, having a weak family connection does not have the same implications in the first country as it does in the second one. More specifically, whereas in the first country, lack of familial network might be related to loneliness, lack of general social capital and isolation, in the second country, the quality of the family network might not be informative at all about other aspects of a person's social or psychological well-being. The two countries thus differ in how different aspects or measures of psychological well-being are interrelated, which is essentially a difference in data structures. And therefore, comparing the two countries in these measures is not a meaningful endeavor.

In this section we will use the PCADSC tools to unveil such differences between Denmark and Bulgaria. These two countries have previously been highlighted to be very different in terms of what defines happiness (\hl{REF: ESS6 Topline Results Series 5}). This might be the result of large, cultural and socioeconomic differences between the two countries. Therefore, a successful method for data comparisons should be able to detect these differences by looking at data on psychological well-being from these two countries.

We also compare the Danish data with Swedish data in order to illustrate that the PCADSC tools actually do have some discriminatory power. Denmark and Sweden are both Scandinavian countries and are often deemed very similar in terms of culture and history. Therefore, we expect fundamental concepts such as psychological well-being to similar across these two countries.

%As Denmark has repeatedly been awarded with the title of "the happiest country", we have included this country

%Specifically, we take on a critical perspective on

%, a very large dataset that is freely available online at \url{www.europeansocialsurvey.org}. As the name suggests, the data comes from a survey that was conducted primarily in Europe aiming to collect information about the social conditions of the citizens \hl{reference?}. As with all international (or, simply, multi-center) studies, one might be concerned about whether or not the data from different countries can readily be combined. This is the question we will address using PCADSC in the current section.

@@

\subsection{Data}
%\begin{figure}
%\center
%\includegraphics[scale=1]{surveyTable.png}
%\caption{\hl{Make a new table here}}
%\label{surveyTable}
%\end{figure}

%\begin{table}[t]
%\centering
%\begin{tabular}{cl}
%Wellbeing dimension & ESS item \\
%\multirow{2}{*}{Evaluative wellbeing} & X \\ & Y \\ 
%\end{tabular}
%\end{table}

\begin{table}[t]
\centering
\begin{tabular}{lccccccccc}
  \hline
  & \multicolumn{3}{c}{Denmark} & \multicolumn{3}{c}{Bulgaria} & \multicolumn{3}{c}{Sweden} \\
 & $Q_1$ & $M$ & $Q_3$ \quad & $Q_1$ & $M$ & $Q_3$ \quad & $Q_1$ & $M$ & $Q_3$ \\ 
  \hline
 Evaluative wellbeing & 8.00 & 8.75 & 9.50 & 3.50 & 5.00 & 7.00 & 7.00 & 8.00 & 9.00 \\ 
 Emotional wellbeing & 7.22 & 8.33 & 8.89 & 5.00 & 6.67 & 7.78 & 6.67 & 7.78 & 8.89 \\ 
Functioning & 6.93 & 7.57 & 8.21 & 5.50 & 6.68 & 7.68 & 6.39 & 7.04 & 7.68 \\ 
Vitality & 6.67 & 7.50 & 8.33 & 5.83 & 7.50 & 8.33 & 6.67 & 7.50 & 9.17 \\ 
 Community wellbeing & 5.83 & 6.77 & 7.57 & 3.70 & 4.67 & 5.70 & 5.66 & 6.57 & 7.37 \\ 
Supportive relationships & 7.42 & 8.25 & 8.92 & 6.17 & 7.25 & 8.08 & 7.42 & 8.25 & 8.75 \\ 
   \hline
\end{tabular}
\caption{The 1st quartile, the median and the third quartile of the distrubutions of each of the six dimensions of psychological well-being, stratified by country. Note that the scales are constructed such that they all run from 0-10.}
\label{tableDistr}
\end{table}

The ESS 2012 data contains a total of 626 variables collected from 54673 citizens in 29 countries. Here, we will only work with a subset of 35 questionnaire items that are all related to psychological well-being. These 35 items can be divided into 6 distinct scales, namely \textit{Evaluative wellbeing}, \textit{Emotional wellbeing}, \textit{Functioning}, \textit{Vitality}, \textit{Community wellbeing} and \textit{Supportive relationships}. More details on these scales is to be found in (\hl{ref: ESS6 Topline Results Series 5}). We represent each of these scales by a single variable, which is calculated as the average score within the items related to that variable (\hl{ref to someone that says that is sensible?}) and scaled such that it takes a value between 0 and 10. For simplicity, we use only complete cases for this construction and thus exclude all participants that did not answer all the 35 questionnaire items used below. This gives us $n_{DK} = 1498$ observations in the Danish sample, $n_{BG} = 1798$ observations in the Bulgarian sample and $n_{SE} = 1736$. Table \ref{tableDistr} summarizes the marginal distributions of the six dimensions of psychological well-being, stratified by country.

\subsection{Comparing Denmark and Bulgaria}
\begin{figure}
\center
\includegraphics[scale = 0.7]{essDKBGce.pdf}
\includegraphics[scale = 0.7]{essDKBGhair.pdf}
\caption{The CE plot (top) and the angle plot (bottom) resulting from comparing Bulgarian and Danish data on psychosocial well-being. Dataset 1 refers to the Bulgarian subsample, while Dataset 2 is the Danish data. The CE plot is annotated with the $p$-values of the Kolmogorov-Smirnov and the Cram\'er-von Mises tests of the assumption of no difference in data structures. In the angle plot, the blue arrows show the principal components of the Bulgarian dataset decomposed in the coordinate system of the principal components of the Danish dataset, while the red arrows illustrate the reverse.}
\label{plotBG.cehair}
\end{figure}

\begin{figure}
\center
\includegraphics[scale = 0.7]{essDKBGpancake234.pdf}
\caption{\hl{something}}
\label{plotBG.pancake}
\end{figure}

%\begin{figure}
%\center
%\includegraphics[scale = 0.7]{essDKBGWallyPCADSC234.pdf}
%\caption{\hl{something.}}
%\label{plotBG.wally}
%\end{figure}

Figure \ref{plotBG.cehair} presents the CE plot and the angle plot obtained from comparing the Danish and Bulgarian psychological well-being scales. The CE plot show a remarkable degree of lacking comparability: The cumulative difference in the eigenvalue by far exceeds what could come about randomly if there really was no difference in the data structure. This is also confirmed by the Kolmogorov-Smirnov and the Cram\'er-von Mises tests, which both result in $p$-values that are virtually zero.

Moving on to the angle plot, we find that the differences are primarily to be found in the second, third and fourth principal components (PCs): The blue arrows visualize the decomposition of the principal components for the first dataset in the coordinate system of the second dataset. We see that PC2 also loads on PC3, that PC3 also loads on PC4, and that PC4 also loads on PC2 and PC3. The red arrows visualize the decomposition of the principal components for the second dataset in the coordinate system of the first dataset. Here, we see that PC2 also loads on PC4, that PC3 also loads on PC2 and PC4, and that PC4 also loads on PC3. Thus, if we wish to understand why differences in the data structures occur, an inspection of the loadings of components 2, 3 and 4 might be informative.

 The chroma blot in Figure \ref{plotBG.pancake} allows us to look closer into these components. Here, we find that the relative importance of the \textit{Community wellbeing} and \textit{Supportive relationships} scales is much larger in the Bulgarian sample than in the Danish. In the Danish data, on the other hand, we find that \textit{Vitality} and \textit{Emotional well-being} seem to play bigger roles, as they appear with larger loadings in more high-ranking components in this sample, relative to the Bulgarian.

All in all, we find that psychological well-being does not seem to be the same concept in Bulgaria as in Denmark. The two countries disagree both in how many dimensions are needed to capture the most important parts of the concept (as illustrated by the differences in eigenvalue) and in how these dimensions are then weighted among the 6 scales (as illustrated by the angle and chroma plots). In Bulgaria, interpersonal features seem to be more informative of psychological wellbeing, whereas in Denmark, individual characteristic play a relatively bigger role. Thus, the datasets are fundamentally different and that we should therefore be wary about combining them in a joint analysis, which was also the conclusion of the ESS authors, though based on country-level aggregated statistics. \\

\hl{I have dropped the Wally-plots for now, as I'm not really sure they are that useful.}

% While the first principal component, which is responsible for explaining 50-60 \% of the variance in the data, is very similar for the two countries, we see quite large differences in the remaining components. In the second component, we see that the two countries disagree in the relative importance of the scales \textit{Community wellbeing} and \textit{Vitality}. In the third and fourth components, general disagreement  is found. All in all, components 2-4, representing almost half of the variability in the data, are not very similar across the two countries. Moreover, the two subsets of the data also differ with respect to how much variance is explained by each component, and the difference is particularly big for the first component. This component has approximately 15 \%  more explanatory power in the Bulgarian subsample than it does in the Danish.

%{\color{red}
%Figure \ref{plotESSHairplot} shows the hairplot. The blue arrows visualize the decomposition of the principal components for the %first dataset in the coordinate system of the second dataset. We see that PC2 also loads on PC3, that PC3 also loads on PC4, and %that PC4 also loads on PC2 and PC3. The red arrows visualize the decomposition of the principal components for the second dataset %in the coordinate system of the first dataset. We see that PC2 also loads on PC4, that PC3 also loads on PC2 and PC4, and that %PC4 also loads on PC3. For PC1, PC5 and PC6 the main difference is in the size of the variation component.
%}

%But did we really illustrate a data structure difference due to country differences or did we just illustrate the variability of the results of the PCADSC method? In order to investigate this further, we look at a so-called \textit{Wally plot} (\hl{ref: Claus Ekstrøm}). In this plot, we compare the results of PCADSC conducted with grouping by country with several random, but similar grouping variables. Specifically, we produce 7 PCADSC plots where the country variable was replaced by a randomly generated variable that divides the observations into two groups of the same sizes as the country samples. The results are illustrated in Figure \ref{plotESSPCADSCWally}. Here, we see that the differences in the second component from the original PCADSC results are not matched in any of the randomly grouped PCADSC runs. In fact, the 7 runs are remarkably similar, thereby illustrating that PCADSC seems to be very robust with respect to random groupings: The signal in the data is not blurred by the random subdivisions. When it comes to the differences in the third component for the two groups, we find much larger variability in the 7 random runs. \hl{more comments here... Wait until we are sure exactly what we think about the results and what other PCA-based methods, we will do before/after. Particularly, how do we deal with eigen value differences?}

\subsection{Comparing Denmark and Sweden}
\begin{figure}
\center
\includegraphics[scale = 0.7]{essDKSEce.pdf}
\includegraphics[scale = 0.7]{essDKSEhair.pdf}
\caption{A CE (top) and a hair (bottom) plot for comparing the Danish (Dataset 1) and the Swedish (Dataset 2) psychological well-being data. The blue arrows show the principal components of the Danish dataset decomposed in the coordinate system of the principal components of the Swedish dataset, and the red arrows illustrate the reverse.}
\label{plotSE.cehair}
\end{figure}

\begin{figure}
\center
\includegraphics[scale = 0.7]{essDKSEpancake.pdf}
\caption{A chroma plot for comparing the loading patterns of the Danish and the Swedish subsamples. Note that the bars for each component is annotated with its cumulative variance score, that is, how much variance can be explained by having information of this and the preceding components.}
\label{plotSE.pancake}
\end{figure}

%\begin{figure}
%\center
%\includegraphics[scale = 0.7]{essDKSEWallyPCADSC.pdf}
%\caption{\hl{something.}}
%\label{plotSE.wally}
%\end{figure}

We now turn to the comparison of Denmark and Sweden in terms of psychological well-being. Figure \ref{plotSE.cehair} show the CE and angle plots for these two countries. In the CE plot, we now find the cumulative eigenvalue curve to be just within the acceptance region of the null-hypothesis. This is also reflected by the two tests, which now produce $p$-values of $p_{KS} = 0.14$ and $p_{CvM} = 0.09$, respectively, thus accepting the null-hypothesis at the typical 5\% level, but not with overwhelming evidence.

The angle plot from Figure \ref{plotSE.cehair} shows that the two datasets agree very strongly about the relative importance of the six scales in the six PCs, as almost all off-diagonal arrows are practically non-existent. This implies that if one already has e.g. the information held in the first PC from the Danish data, this information is in itself mostly sufficient to describe the first PC of the Swedish data.

Looking at the chroma plot in Figure \ref{plotSE.pancake}, the same tale is told once again: Here, we find remarkably similar loading patterns in the first three components (which are responsible for almost 80 \% of the variance in both datasets), and slight, but increasing, differences in the remaining three components. We therefore conclude that any differences in the data structures of the Danish and the Swedish samples are related to the least important dimensions of the datasets and that these dimensions are only responsible for less than 25 \% of the variance in both datasets. In particular, this means that we can combine and compare the Danish and Swedish datasets in a meaningful way and e.g. conclude using Table \ref{tableDistr} that in general, Danes seem to be somewhat more happy than Swedes, and in particular, the unhappy people in Denmark (represented by the 1st quartiles) are generally a lot happier in Denmark than they are in Sweden. A more thorough, statistical investigation could now be put to work on answering \textit{why} this seems to be the case.

\section{Discussion}
\label{sec:discussion}
\hl{
\begin{itemize}
\item Generalizing the results to non-numeric variables?
	\begin{itemize}
		\item Interpretation for binary variables?
		\item Any meaningful way to allow for nominal, categorical variables?
		\item ... ?
	\end{itemize}
\item Generalizing the results to covariance matrices that are not of full rank?
\item Investigate sensitivity towards the sample sizes $n_1$ and $n_2$
\item Limitations: What sorts of problems can never be found using PCADSC? 
\begin{itemize}
\item Differences in scaling, as we standardize all variables
\item More?
\end{itemize}
\end{itemize}
}

%\section{Concluding Remarks}
%\label{sec:conclusion}
